---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Jacob Hogan"
date: "9-25-25"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

Bloodmeal data from Utah demonstrates several species that host West Nile Virus. Host positivity and tests were cross checked and plotted. In this markdown, the hypothesis is that by using the common carrier (House Finces) the areas of increased WNV may be predicted by area of total carrier population. This was used to create other plots that are able to predict positive hosts of WNV. House Finches were were the clear leader in WNV positive hosts. This data is essential in control of the spread of WNV by mosquito controling practices in areas of high populations of House Finches.

# BACKGROUND

West Nile Virus (also known as WNV) is a worldwide virus that has recently traveled to the United States. WNV as recently as August 2003 (Utah Epidemiology 2025), and since then there have been 264 total deaths in the United States since 2003 (CDC 2025). This virus is passed through mosquitos that mediate the virus through infected birds (the natural host of WNV) along with other mammals (IBIS-PH 2025). Symptoms may include fever, headache, body aches, rash, swollen lymph glands (Hopkins Medicine 2025). There has been 1 WNV death in Utah – which occurred in 2023 (Fox13 News 2025).

Viremia is the presence of virus in blood, in this r markdown project House Finches and the viremia levels of WNV is highlighted. This was done to determine their effectiveness as amplifying hosts of WNV through high levels of viremia in mosquito pools (Kumar et al., 2003).

```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Questions  
Which species in Utah are hosting West Nile Virus in Utah?

## Hypothesis 
House Finches are currrently the main host for West Nile Virus in the Utah area.
  
## Prediction
If house finches are the main host of West Nile Virus, the most effective way of controlling the spread of WNV is through targeted attacks of mosquitos in areas that house finches reside. 

# METHODS

Mosquitos were collected using traps, with them then undergoing DNA amplification through DNA digestion and PCR. This data was processed into the document blood_plusWNV_for_BIOL3070.csv. Mosquito data was taken from blood_plusWNV_for_BIOL3070.csv and imported into the R Markdown. The table was organized according to West Nile Virus positive or negative. This was then crossed with a different plot to determine whether west nile positivitiy in House Finches could be predicted.

## First analysis
```{r first-analysis, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```

## Second analysis
```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```

# DISCUSSION

# Horizontal barplots Interpretation:
 This plot shows how common Utah birds are the most likely to be WNV hosts.
 WNV positive and negative samples had differing compositions in species.
 House Finches are the most common WNV host.
 
 The plot demonstrates that mosquito blood meals of these birds shows a significant increase in WNV contraction.
 
# Statistical model Interpretation:
 The House Finch blood-meal coefficient had a positively statistically significant (p < 0.05).
 
 This shows that as blood meals in a specific area increase, the likelyhood of a positive host test increases aswell.
 
 In short, House Finches are a great indicator of WNV amplification.

# CONCLUSION

In conclusion, the original hypothesis of a common Utah bird being the primary host of WNV and using that bird species to predict the growth of WNV. This hypothesis was proven correct as the barplot and linear model show a correlation between WNV positivity and the host species. To put this finding into our hypothesis, it means that they higher level of host species in an area would prove a greater increase of WNV. This finding is essential to control the spread of WNV as mosquito abatement in those areas of high host species (House Finches).

# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.

3. West Nile virus. Utah Epidemiology. (n.d.). https://epi.utah.gov/west-nile-virus/ 

4. Centers for Disease Control and Prevention. (n.d.). Historic Data (1999-2024). Centers for Disease Control and Prevention. https://www.cdc.gov/west-nile-virus/data-maps/historic-data.html 

5. Health, D. of. (n.d.). Complete health indicator report of west nile virus infections. IBIS. https://ibis.utah.gov/ibisph-view/indicator/complete_profile/WNVInf.html 

6. Michael Martin, S. M. (2025, September 3). Utah officials confirm 8 West Nile virus cases, including Fatal case. FOX 13 News Utah (KSTU). https://www.fox13now.com/news/local-news/northern-utah/utah-county-officials-confirm-5-west-nile-virus-cases-4-with-neuroinvasive-symptoms 



